---
title: "SMS Text Mining - Spam vs Legitimate"
author: "Olivia Azevedo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: true
---

## Overview

Demonstrates tidying text data and performing text processing via lexicon-based sentiment analysis. The corpus used for this assignment is a collection of SMS messages tagged as spam or legitimate. The files contain one message per line. Each line is composed by two columns:

  * spam_flag: contains the label (ham or spam)
  * text: contains the raw text.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidytext)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(textdata)
library(groupdata2)
library(forcats)

spam_colors = c("Spam" = "tomato", "Legitimate" = "steelblue")
```


<hr style="border:1px solid #2d3c52">


## Load Data
```{r}
data = read.csv('https://raw.githubusercontent.com/okazevedo90/DATA607_GitHub/refs/heads/main/weekly_assignments/text_mining/spam.csv')
slang_ref = read.csv('https://raw.githubusercontent.com/okazevedo90/DATA607_GitHub/refs/heads/main/weekly_assignments/text_mining/slang.csv')
```

### Create Stop Word DataFrame
```{r}
sms_stop_words = c('lor')
lexi = rep(c('custom'), length(sms_stop_words))

custom_stop_words = bind_rows(
  tibble(word = sms_stop_words, lexicon = lexi),
  stop_words)
```


<hr style="border:1px solid #2d3c52">


## Clean and Tidy Text Data

* Update `spam_flag` column values
* Remove html characters
* Unnest and Tidy SMS text by making each token or word one row
* Join and replace common text slang / abbreviations with english meaning
* Unnest SMS words again to account for any slang replacements that are more than one word
```{r}
data$text = iconv(data$text, to = "ASCII", sub = "")


df = data |>
  mutate(
    id = row_number(),
    spam_flag = ifelse(spam_flag == 'ham', 'Legitimate', 'Spam'),
    # Remove html characters
    text = str_replace_all(str_replace_all(
      str_replace_all(
        str_replace_all(
          text, '&amp', ''), '&gt', ''), '&lt', ''), "’", "'")
        ) |>
  unnest_tokens(slang, text) |>
  left_join(slang_ref, by = join_by(slang)) |>
  mutate(word = ifelse(is.na(word), slang, word)) |>
  unnest_tokens(word, word)

# df = downsample(df, 'spam_flag')
```


<hr style="border:1px solid #2d3c52">


## Sentiment and Word Frequency Analysis - Spam vs Legitimate SMS

Identify Most Common Words
```{r, warning=FALSE}
df |>
  anti_join(custom_stop_words, by = join_by(word)) |>
  group_by(spam_flag) |>
  count(word, sort = TRUE) |>
  slice_max(n, n = 10) %>% 
  mutate(word = reorder(word, n)) |>
  ggplot(aes(n, word, fill = spam_flag)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~spam_flag, scales = "free_y") +
  scale_fill_manual(values = spam_colors) +
  labs(
    title = 'Most Common Spam vs Legitimate SMS Words',
    x = 'Count', y = NULL
    ) + 
  theme_classic()

df |>
  inner_join(custom_stop_words, by = join_by(word)) |>
  group_by(spam_flag) |>
  count(word, sort = TRUE) |>
  slice_max(n, n = 10) %>%
  mutate(word = reorder(word, n)) |>
  ggplot(aes(n, word, fill = spam_flag)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = spam_colors) +
  facet_wrap(~spam_flag, scales = "free_y") +
  labs(
    title = 'Most Common Spam vs Legitimate SMS STOP Words',
    x = 'Count', y = NULL
    ) + 
  theme_classic()
```


Identify Most Common Sentiment Text Categories
```{r, warning=FALSE}
df |>
  anti_join(custom_stop_words, by = join_by(word)) |>
  inner_join(get_sentiments("nrc"), by = join_by(word)) %>%
  group_by(spam_flag) |>
  mutate(sf_n = n()) |>
  ungroup() |>
  group_by(spam_flag, sentiment) |>
  summarise(n = n() / sf_n, .groups='keep') |>
  distinct() |>
  ggplot(aes(reorder(sentiment, n), n, fill = spam_flag)) +
  geom_col(position='dodge') +
  scale_fill_manual(values = spam_colors) +
  coord_flip() +
  labs(title = "SMS NRC Lexicon Sentiment Category Word Distribution
       Spam vs Legitimate",
       y = 'Percent SMS Type', x = 'Sentiment Category', fill = 'SMS Type') +
  theme_classic()
```


Identify Distribution of Overall SMS Net Sentiment Score
```{r}
df |>
  anti_join(custom_stop_words, by = join_by(word)) |>
  inner_join(get_sentiments("bing"), by = join_by(word)) |>
  count(spam_flag, id, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |> 
  mutate(sentiment = positive - negative) |>
  ggplot(aes(sentiment, fill = spam_flag)) +
  geom_histogram(bins = 25, show.legend = FALSE, color = 'black') +
  scale_fill_manual(values = spam_colors) +
  facet_wrap(~spam_flag) +
  labs(
    title = 'SMS Net Sentiment Bing Score Distribution - Spam vs Legitimate',
    x = 'Net Sentiment Bing Score (positive - negative)', y = 'Count'
    ) + 
  theme_classic()

```


Identify Differences in SMS Lexicon Sentiment Scores 
```{r, warning=FALSE}
afinn = df |>
  anti_join(custom_stop_words, by = join_by(word)) |>
  inner_join(get_sentiments("afinn"), by = join_by(word)) %>% 
  group_by(id, spam_flag) %>% 
  summarise(sentiment = sum(value), .groups='keep') %>% 
  mutate(method = "AFINN")

bing_and_nrc = bind_rows(
  df |>
    anti_join(custom_stop_words, by = join_by(word)) |>
    inner_join(get_sentiments("bing"), by = join_by(word)) %>%
    mutate(method = "Bing et al."),
  df |>
    anti_join(custom_stop_words, by = join_by(word)) |>
    inner_join(get_sentiments("nrc") %>% 
                 filter(sentiment %in% c("positive", 
                                         "negative")),
               by = join_by(word)
    ) %>%
    mutate(method = "NRC")) %>%
  count(spam_flag, method, id, sentiment) %>%
  pivot_wider(names_from = sentiment,
              values_from = n,
              values_fill = 0) %>% 
  mutate(sentiment = positive - negative)

bind_rows(afinn, 
          bing_and_nrc) %>%
  ggplot(aes(sentiment, method, fill=method)) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(~spam_flag) +
  labs(
    title = 'SMS Sentiment Score by Lexicon - Spam vs Legitimate',
    x = 'Sentiment Score', y = 'Lexicon'
    ) + 
  theme_classic()
```


Identify Words Contributing to Highest Positive and Negative Sentiment Scores
```{r}
df |>
  anti_join(custom_stop_words, by = join_by(word)) |>
  inner_join(get_sentiments("bing"), by = join_by(word)) %>%
  count(spam_flag, word, sentiment, sort = TRUE) %>%
  ungroup() %>%
  group_by(spam_flag, sentiment) %>%
  slice_max(n, n = 5) %>% 
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col() +
  scale_fill_manual(values = c('darkorange', 'forestgreen')) +
  facet_wrap(~spam_flag, scales = "free_y") +
  labs(x = "Contribution to sentiment",
       y = NULL) +
  theme_classic()
```


Identify Most Important Words per SMS Type within all SMS text (tf-idf)
```{r}
sms_words <- df |>
  anti_join(custom_stop_words, by = join_by(word)) |>
  count(spam_flag, word, sort = TRUE)

total_words <- sms_words %>% 
  group_by(spam_flag) %>% 
  summarize(total = sum(n))

sms_words <- left_join(sms_words, total_words, by = join_by(spam_flag))

# ggplot(sms_words, aes(n/total, fill = spam_flag)) +
#   geom_histogram(show.legend = FALSE) +
#   xlim(NA, 0.0009) +
#   facet_wrap(~spam_flag, ncol = 2, scales = "free_y")

freq_by_rank <- sms_words %>% 
  group_by(spam_flag) %>% 
  mutate(rank = row_number(), 
         term_frequency = n/total) %>%
  ungroup()

# freq_by_rank %>% 
#   ggplot(aes(rank, term_frequency, color = spam_flag)) + 
#   geom_line(linewidth = 1.1, alpha = 0.8, show.legend = FALSE) + 
#   scale_x_log10() +
#   scale_y_log10()

sms_tf_idf <- sms_words %>%
  bind_tf_idf(word, spam_flag, n) %>%
  select(-total) %>%
  arrange(desc(tf_idf))

sms_tf_idf %>%
  group_by(spam_flag) %>%
  slice_max(tf_idf, n = 15) %>%
  ungroup() %>%
  ggplot(aes(tf_idf, fct_reorder(word, tf_idf), fill = spam_flag)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = spam_colors) +
  facet_wrap(~spam_flag, ncol = 2, scales = "free_y") +
  labs(
    title = 'SMS tf-idf Important Words - Spam vs Legitimate',
    x = 'tf-idf', y = NULL
    ) + 
  theme_classic()
```

