---
title: "Class Repo New Submission"
author: "Barakat Adigun"
date: "2025-04-08"
output: html_document
---

## In this Repo, I will be sharing my work on Project 1

## Project 1

## In this project, you’re given a text file with chess tournament results where the information has some structure. Your job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players: Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents. For the first player, the information would be: Gary Hua, ON, 6.0, 1794, 1605. 1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.

```{r setup, include=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
```

## Loading Data

```{r}
# Load Raw Text file Data from github

file_path <- 'https://raw.githubusercontent.com/Badigun/Data-607-Assignments/refs/heads/main/chess%20tournament%20file.txt'

tournament_data <- read_lines(file_path)

# View the first few lines of the Data
head(tournament_data)
```

```{r}
tail(tournament_data)
```

## Edit Player information in the Data
I observe that the data follows a specific pattern. The first four lines contain non-data information, so they are excluded from the dataset. After that, the player details and game statistics appear in a repeating sequence every three lines. The data is then organize into two matrices, one for player information and the other for their game statistics, to structure it like a more conventional dataset.

Each player has two lines in the file, which includes;

Line 1: Name, total points, and opponent numbers.

Line 2: State and pre-rating.

```{r}
# Name, total points, and opponent numbers
edit_tournament_data <- matrix(unlist(tournament_data), byrow=TRUE)

m1 <- edit_tournament_data[seq(5,length(edit_tournament_data),3)]
head(m1)
```

Another way to extract player information for Line 1: Name, total points, and opponent numbers

```{r}
player_lines <- tournament_data[grep("^\\s*\\d+", tournament_data)]

# Identify lines that contain player information 
player_lines <- tournament_data[grep("^\\s*\\d+", tournament_data)]

# Extract Player's Name, Points, and Opponents
players_info <- lapply(player_lines, function(line) {
  elements <- unlist(strsplit(line, "\\|"))
  elements <- trimws(elements)  # Remove leading and trailing spaces
  
  # Extract fields
  name <- elements[2]  # Player Name
  total_points <- as.numeric(elements[3])  # Total Points
  opponents <- unlist(str_extract_all(elements[6:length(elements)], "\\d+"))  # Opponent Numbers
  
  return(list(name, total_points, opponents))
})

# Convert to DataFrame
players_df <- data.frame(
  Name = sapply(players_info, `[[`, 1),
  Points = sapply(players_info, `[[`, 2),
  Opponents = sapply(players_info, function(x) paste(x[[3]], collapse = ",")), # Convert list to string
  stringsAsFactors = FALSE
)
head(players_df)
```

This code is to extract Line 2: State and pre-rating

```{r}
m2 <- edit_tournament_data[seq(6,length(edit_tournament_data),3)]
head(m2)
```

Another way to extract Line 2: State and pre-rating

```{r}
# Find the second line for each player's entry (contains state & rating)
rating_lines <- tournament_data[grep("R:\\s*\\d+", tournament_data)]

# Extract State and Pre-Rating
ratings_info <- lapply(rating_lines, function(line) {
  elements <- unlist(strsplit(line, "\\|"))
  elements <- trimws(elements)

  state <- substr(elements[1], 1, 2)  # Extract first two letters as state
  pre_rating <- as.numeric(str_extract(elements[2], "(?<=R:\\s)\\d+"))  # Extract pre-rating

  return(list(state, pre_rating))
})

# Add to DataFrame
players_df$State <- sapply(ratings_info, `[[`, 1)
players_df$Pre_Rating <- sapply(ratings_info, `[[`, 2)
```

## Capturing The Data
Since the Data is more organized, capturing of the data featured can be done

```{r}
# Convert m1 and m2 to character vectors 
m1 <- as.character(m1)
m2 <- as.character(m2)

# matching first numbers
ID <- as.numeric(str_extract(m1, '\\d+'))

# matching the first combination of a letter, any amount of characters and "|"
Name <- str_extract(m1, '[A-z].{1,32}') 

# extracting the name
Name <- str_trim(str_extract(Name, '.+\\s{2,}'))

# matching the first two letters (state) in the second matrix 
State <- str_extract(m2, '[A-Z]{2}') 

# matching at least 1 number, a period, and 1 number
Total_Points <- as.numeric(str_extract(m1, '\\d+\\.\\d'))

# matching the combination of "R", any characters and "-"
PreRating <- str_extract(m2, 'R:.{8,}-')

# matching first 4 numbers
PreRating <- as.numeric(str_extract(PreRating, '\\d{1,4}'))

# matching all combinations of 1 letter 2 spaces and any numbers
Rounds <- str_extract_all(m1, '[A-Z]\\s{2,}\\d+')

# matching numbers
Rounds <- str_extract_all(Rounds, '\\d+')
```

## Compute Average Opponent Pre-Rating

Compute the average pre-rating of opponents using the vectors from the previous step
```{r}
AvgOpp_Rating <- c()

for(i in c(1:length(Rounds))){
  AvgOpp_Rating[i] <- round(mean(PreRating[as.numeric(Rounds[[i]])]),0)
}
AvgOpp_Rating
```

```{r}
Project1_Data607 <- data.frame(ID,Name,State,Total_Points,PreRating,AvgOpp_Rating)

head(Project1_Data607)
```

## Writing/Saving Data to CSV

```{r}
write_csv(Project1_Data607, "ChessTournament_results.csv")
```


____