---
title: "Assignment9_DATA607"
author: "Aaliyah John-Harry"
date: "2025-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(dplyr)
library(openintro)
```


```{r}
data("fastfood", package='openintro')

mcdonalds <- fastfood %>%
  filter(restaurant == "Mcdonalds")
dairy_queen <- fastfood %>%
  filter(restaurant == "Dairy Queen")

fastfood
head(mcdonalds)
head(dairy_queen)
```
- What is the probability that a randomly chosen McDonald's product has more than 1000 total calories?
- What is the probability that a randomly chosen Dairy Queen product has more than 25 total carbs?

```{r}
mcdonalds %>% 
  filter(calories > 1000) %>%
  summarise(percent = n() / nrow(mcdonalds))

```
```{r}
dairy_queen %>% 
  filter(total_carb > 25) %>%
  summarise(percent = n() / nrow(dairy_queen))
```
How can we classify high calorie and low calorie products sold by McDonald's? 
How many are there of each?
```{r}
mcdonalds <- mcdonalds %>%
  mutate(calorie_level = ifelse(calories > 1000, "High", "Low"))
head(mcdonalds)
```

```{r}
mcdonalds %>%
  group_by(calorie_level) %>%
  summarise(count = n())
```
Top 10 highest calorie items at McDonald's
```{r}
mcdonalds %>%
  arrange(desc(calories)) %>%
  head(10)
```
How many unique items are sold by each fast food restaurant?
```{r}
fastfood %>%
  group_by(restaurant) %>%
  summarise(unique_items = n_distinct(item))
```


#Daniel Hanasab changes : simple bar chart showing how many “High” and “Low” calorie items McDonald’s sells:
library(ggplot2)

ggplot(mcdonalds, aes(x = calorie_level, fill = calorie_level)) +
  geom_bar() +
  labs(title = "McDonald's Products by Calorie Level",
       x = "Calorie Category",
       y = "Number of Items") +
  theme_minimal()

