---
title: "Project_Proposal_WF"
author: "William Forero"
output:
  pdf_document: default
  html_document: default
---
---
title: DATA 606 Data Project Proposal
author: 
---

### Data Preparation

```{r}
library(tidyverse)
```

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
raw_ipeds <- read_csv("https://raw.githubusercontent.com/william-forero/Data-607/refs/heads/main/2223_IPEDS_CUSTOM_DATA_FILE.csv")
```

```{r}
glimpse(raw_ipeds)
```

```{r}
head(raw_ipeds)
```

```{r}
summary(raw_ipeds)
```

```{r}
ipeds_renamed <- raw_ipeds |>
  rename(
    unit_id = 'unitid',
    inst_name = 'institution name',
    year = 'year',
    state = 'HD2023.FIPS state code',
    inst_2_4_yr = 'HD2023.Sector of institution',
    ft_ret_rate = 'EF2023D.Full-time retention rate, 2023',
    tot_nbr_ugrd_fa = 'SFA2223.Total number of undergraduates - financial aid cohort',
    tot_nbr_ugrd = 'SFA2223.Total number of degree/certificate-seeking undergraduate students',
    nbr_ugrd_degr_awarded = 'SFA2223.Number of degree/certificate-seeking undergraduate students awarded federal, state, local, institutional or other sources of grant aid',
    perc_ugrd_degr_awarded = 'SFA2223.Percent of degree/certificate-seeking undergraduate students awarded federal, state, local, institutional or other sources of grant aid',
    tot_aid_degr_awarded = 'SFA2223.Total amount of federal, state, local, institutional or other sources of grant aid awarded to degree/certificate-seeking undergraduate students',
    avg_aid_degr_awarded = 'SFA2223.Average amount of federal, state, local, institutional or other sources of grant aid awarded to degree/certificate-seeking undergraduate students',
    tot_grad_rt = 'DRVGR2023.Graduation rate, total cohort',
    grad_rt_4_yr = 'DRVGR2023.Graduation rate - Bachelor degree within 4 years, total',
    grad_rt_5_yr = 'DRVGR2023.Graduation rate - Bachelor degree within 5 years, total',
    grad_rt_6_yr = 'DRVGR2023.Graduation rate - Bachelor degree within 6 years, total',
  ) |>
  select(unit_id,inst_name,year,state,inst_2_4_yr,ft_ret_rate,tot_nbr_ugrd_fa,tot_nbr_ugrd,nbr_ugrd_degr_awarded,perc_ugrd_degr_awarded,tot_aid_degr_awarded,tot_aid_degr_awarded,avg_aid_degr_awarded,tot_grad_rt,grad_rt_4_yr,grad_rt_5_yr,grad_rt_5_yr,grad_rt_6_yr)

head(ipeds_renamed)
```


### Research question 

Is there a statistically significant relationship between the average amount of financial aid awarded and student success outcomes (graduation rates) at undergraduate degree-granting institutions in the United States?


### Cases 

The cases are represented by colleges included in the dataset that was pulled from IPEDS. There are a total of 1,563 distinct colleges included in the dataset.


### Data collection 

The data is pulled from the Integrated PostSecondary Education Data System (IPEDS) using their custom data file tool. Information was pulled for degree granting 2 and 4 year public institutions within the United Stated for the year 2022-2023. 


### Type of study 

This is an observational study.


### Data Source 

https://nces.ed.gov/ipeds/datacenter/CDSPreview.aspx?sid=080fa437-d472-40b0-8585-ecaf99adfab5&rtid=5 


### Describe your variables?

My dependent variable will be the graduation rate. 

My variables are predominantly quantitative including sums of aid awarded for degree seeking undergraduate students, counts of students awarded, and graduation rates.

The qualitative variables are those that specify if a college is a 2 year or 4 year college and the state.

### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r Nbr of colleges included}
ipeds_renamed |>
  summarise(num_colleges = n())
```


```{r Nbr of colleges included by state}
ipeds_renamed |>
  filter(!is.na(state)) |>
  group_by(state) |>
  summarise(num_colleges = n(),
            nbr_stdnt = sum(tot_nbr_ugrd, na.rm = TRUE),
            nbr_stdnt_awd = sum(nbr_ugrd_degr_awarded, na.rm = TRUE),
            tot_aid_awarded = sum(tot_aid_degr_awarded, na.rm = TRUE),
            avg_grad_rate = round(mean(tot_grad_rt, na.rm = TRUE),2))
```

The graduation rate is slightly right skewed and has a natural floor as the graduation rate cannot be below 0%. 

```{r Visualization on grad rate}
ipeds_renamed |>
  filter(!is.na(tot_grad_rt)) |>
  ggplot(aes(x = tot_grad_rt)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Distribution of Graduation Rates", x = "Graduation Rate (%)", y = "Number of Colleges")
```


```{r scatterplot for graduation rate and financial aid awarded}
ipeds_renamed |>
  filter(!is.na(tot_grad_rt) & !is.na(avg_aid_degr_awarded)) |>
  ggplot(aes(x = avg_aid_degr_awarded, y = tot_grad_rt)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Grant Aid vs. Graduation Rate", x = "Avg Aid Awarded", y = "Graduation Rate")
```


