---
title: "Class Repo New Submission"
author: "Gillian McGovern, addition by Maxfield Raynolds and Aaliyah John-Harry"
date: "2025-04-07"
output: html_document
---

## Overview

The tidyverse function I am going to look at is `between` from dplyr.

## Load the Libraries

```{r}
library(tidyverse)
library(nycflights13)
data(flights)
```


## Between Function

I chose `between` because it's a way more efficient way to do x >= left & x <= right which I tend to do a lot.

Let's test out the between function using `nycflights`:

```{r}
filter(flights, between(distance, 800, 1500))
```

Now we can easily see the filtered distance flights without typing out distance >= 800 & x <= 1500.

---
Addition by Maxfield Raynolds:

## Inverse of between!

In the interest of using between, we could also use the inverse of between by including an exclamation point!

Using the !between() we can filter all of the flights that are less than 800 or more than 1500 miles.

```{r using !between}
filter(flights, !between(distance, 800, 1500))
```

Addition by Aaliyah John-Harry:

# Add a new distance category column

I added a column that categorizes the flight distance as short, medium and long making it easier to summarize and visualize.

```{r}
flights_updated <- flights %>%
  mutate(distance_category = case_when(
    between(distance, 800, 1500) ~ "medium (800-1500 mi)",
    distance < 800              ~ "short (< 800 mi)",
    distance > 1500             ~ "long (> 1500 mi)"
  ))
```

# Count flights in each category
```{r}
flight_counts <- flights_updated %>%
  count(distance_category)
```