---
title: "Election Forecast 2022"
author: "Alina Vikhnevich"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
library(readxl)

# Load dataset
senate_national_toplines_2022 <- read_excel("election-forecasts-2022/senate_national_toplines_2022.xls")

# Verify dataset is loaded correctly
if (!exists("senate_national_toplines_2022")) {
  stop("Error: Dataset 'senate_national_toplines_2022' not found.")
}

# Data Cleaning & Transformation
election_df <- senate_national_toplines_2022 %>%
  select(forecastdate, branch, chamber_Dparty, chamber_Rparty, popvote_margin) %>%
  rename(
    Date = forecastdate,
    Branch = branch,
    Democratic_Seats = chamber_Dparty,
    Republican_Seats = chamber_Rparty,
    Popular_Vote_Margin = popvote_margin
  )

# Convert Date to proper format
election_df$Date <- as.Date(election_df$Date, format="%Y-%m-%d")
election_df <- election_df %>% drop_na(Date)
```
## Introduction

The 2022 FiveThirtyEight Election Forecast provides predictions for the U.S. Senate races based on polling, historical data, and statistical models.
This dataset offers insights into the expected number of seats for Democratic and Republican candidates and the popular vote margin over time.

In this analysis, we aim to clean, transform, and visualize the dataset to better understand election trends.
By exploring seat projections and vote margin fluctuations, we can observe how public opinion shifted leading up to the election.
---

## Data Visualization

The following plots visualize key election trends:

The first plot shows the forecasted number of Senate seats for the Democratic and Republican parties over time.

```{r, echo=TRUE, fig.width=8, fig.height=5}
ggplot(election_df, aes(x=Date)) +
  geom_line(aes(y=Democratic_Seats, color="Democratic")) +
  geom_line(aes(y=Republican_Seats, color="Republican")) +
  theme_minimal() +
  labs(
    title="Senate Seats Forecast Over Time",
    x="Date",
    y="Number of Seats"
  ) +
  scale_color_manual(values = c("Democratic" = "blue", "Republican" = "red"))
```

The second plot illustrates how the predicted popular vote margin fluctuated throughout the election period.

```{r, include=TRUE, fig.width=8, fig.height=5}
ggplot(election_df, aes(x=Date, y=Popular_Vote_Margin, color=Branch)) +
  geom_line() +
  theme_minimal() +
  labs(
    title="Popular Vote Margin Over Time",
    x="Date",
    y="Popular Vote Margin"
  )
```
---
## Conclusion

Analyzing the 2022 FiveThirtyEight Election Forecast dataset provided meaningful insights into election trends.
By examining seat forecasts and popular vote margins, we observed how public sentiment shifted over time.

The visualizations highlight how election projections evolved leading up to voting day,
demonstrating the impact of polling trends and external influences.

This project was a great opportunity to work with real-world election data,
offering a deeper understanding of how predictive models inform political analysis.