---
title: "LAB10_1"
author: "Farhod Ibragimov"
date: "2025-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(dplyr)
library(httr)
library(jsonlite)
library(tidyr)
library(purrr)
```

## Description

This is my ongoing research project I am working on.

In this project I am using my several favorite libraries to download, make data tidy, and joining in one dataset. As I mentioned before, this is my ongoing not finished project. I am planning to add other datasets and make a research how inflation affects median incomes.

## Downloading data

Here I wrote a function that:

-   Connects to my GitHub repo

-   Finds all files in specified folder

-   Converts links to raw download format

Then I plug in my GitHub username, repo name, folder path, and branch name to get the list of data file links. It prints out all the file URLs.

```{r}

get_github_raw_urls <- function(user, repo, path, branch = "main") {
  api_url <- paste0("https://api.github.com/repos/", user, "/", repo, "/contents/", path, "?ref=", branch)
  
  response <- GET(api_url)
  
  contents <- fromJSON(content(response, "text", encoding = "UTF-8"))
  
  files <- contents[contents$type == "file", "path"]
  
  urls <- paste0("https://raw.githubusercontent.com/", user, "/", repo, "/", branch, "/", files)
  
  return(urls)
}

user <- "farhodibr"
repo <- "CUNY-SPS-MSDS"
path <- "DATA607/LAB10/DATA"  
branch <- "main"

urls <- get_github_raw_urls(user, repo, path, branch)
print(urls)


```

## Read excel file and clean it.

Since all datasets I used are build using the same way and format in excell I wrote a function that:

-   Downloads an Excel file (if it's a URL)

-   Reads a name from a specific cell (B7) to use as the dataframe name.

-   Skips some rows (headers, title rows)

-   Cleans the name (removes punctuation and spaces)

-   Loads the data into R and saves it using that cleaned name

```{r}
read_file_to_df <- function(file_URL, name_cell = "B7", sheet = 1, data_range = NULL) {
 
  if (grepl("^https?://", file_URL)) {
    temp_file <- tempfile(fileext = ".xlsx")
    GET(file_URL, write_disk(temp_file, overwrite = TRUE))
    file_path <- temp_file
  } else {
    file_path <- file_URL
  }

  name_df <- read_excel(file_path, sheet = sheet, range = name_cell, col_names = FALSE) |>
    pull(1) |>
    as.character()

  clean_name <- gsub("[[:punct:]]", "_", name_df)
  clean_name <- gsub("\\s+", "", clean_name)
  name_df <- make.names(clean_name)

  if (is.null(data_range)) {
    data <- read_excel(file_path, sheet = sheet)
  } else {
    data <- read_excel(file_path, sheet = sheet, range = data_range)
  }

  data <- data[-(1:7), ]
  
  data <- data[-1, ]
  colnames(data) <- as.character(data[1, ])
  data <- data[-1, ]
  
  assign(name_df, data, envir = .GlobalEnv)
 
  return(invisible(name_df))
}

```

## Creating a vector with the names of datasets

Here the code loops through each file and creates `created_dataframes`vector containing the names of datasets.

```{r}
created_dataframes <- c()

for (url in urls) {
  df_name <- read_file_to_df(url)
  
  df_name
  
  created_dataframes <- c(created_dataframes, df_name)
}
```

## Tidying dataframes

This part does following:

-   Loop through all the other datasets I loaded:

-   Turns them into long/tidy format

-   Gives each a new column name like `avg_price_of_<dataset_name>`

-   Makes sure the year is an actual number (not text)

-   Saves each tidy version with `_tidy` added to the name

```{r}
for (name in created_dataframes) {
  tidy_df_name <- paste0(name, "_tidy")
  
  df <- get(name)
  
  tidy_df <- df|> 
    pivot_longer(
      cols = -Year,
      names_to = "month",
      values_to = paste0("avg_price_of_", name)
    )
  
  tidy_df$Year <- as.integer(tidy_df$Year)
  assign(tidy_df_name, tidy_df, envir = .GlobalEnv)
  head(tidy_df_name)
}

```

## Creating one dataset by joining all datasets

Since previous function creates tidy datasets by extending dataset's name with `_tidy,`here I create a vector to store tidy datasets names.

```{r}
tidy_names <- paste0(created_dataframes, "_tidy")
```

```{r}
tidy_dfs_list <- lapply(tidy_names, get)
joined_data_tidy <- reduce(tidy_dfs_list, full_join, by = c("Year", "month"))
glimpse(joined_data_tidy)
```
```{r}
print(joined_data_tidy)
```

