# Modification: Instead of using paste0() to manually build an SQL query string 
# (which is error-prone and insecure), it now uses dbExecute() with parameterized 
# queries via placeholders. This prevents SQL injection and handles data types properly.
# Also fixed irrelevant duplicates of value.


---
title: "DATA 607 Week 2 assignment"
author: "Farhod Ibragimov"
date: "2025-02-09"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Build Table

First, I will load necessary libraries:

```{r libraries, message=FALSE, warning=FALSE}
library(DBI)
library(RMySQL)
library(dplyr)
library(ggplot2)
library(googlesheets4)
```

Here is the code to connect and create a table in MySQL database. My credentials are stored in my Windows environment variables.

```{r connect-to-MySQL}
# Credentials to access database from MySQL
db_host <- Sys.getenv("DB_HOST")
db_port <- as.integer(Sys.getenv("DB_PORT"))
db_name <- Sys.getenv("DB_NAME")
db_user <- Sys.getenv("DB_USER")
db_pass <- Sys.getenv("DB_PASS")
print(db_host)

# MySQL database connection
con <- dbConnect(
  drv = RMySQL::MySQL(), 
  dbname = db_name, 
  host = db_host,
  username = db_user,
  password = db_pass
)

query <- paste("CREATE TABLE IF NOT EXISTS Movies_new1 (
              Timestamp DATETIME,
              Name CHAR(100),
              Gladiator2 INT,
              Wolfs INT,
              The_Substance INT,
              Bad_Boys4 INT,
              The_Beekeeper INT,
              Rebel_Ridge INT,
              PRIMARY KEY (Name)
              )", sep = "")

# Create table if it does not exist
dbExecute(con, query)
```

I created a Google Form survey to collect movie reviews. Here the link to the survey:

<https://docs.google.com/forms/d/e/1FAIpQLScraIXSeT6wwZyp4QuA06m1aX5TPUDmccTSz1_a6D5mJWuLBg/viewform>

The form is connected to a Google Sheet, which is publicly accessible and stores survey responses. Here is the link to the Google Sheets:

<https://docs.google.com/spreadsheets/d/1x7uAbOozoNutaFOfbq_X8c33vzSXvqshJxZgu_MfJhI/edit?usp=sharing>

This code chunk pulls the data from Google Sheets:

```{r googlesheets, echo=TRUE, message=FALSE, warning=FALSE}
# URL of the publicly accessible "movie_reviews" Sheet
# Fixed duplicate value
sheet_url <- "https://docs.google.com/spreadsheets/d/1x7uAbOozoNutaFOfbq_X8c33vzSXvqshJxZgu_MfJhI/edit?usp=sharing"
sheet_name <- "Form Responses 1"

# Read the data from the sheet
data <- read_sheet(sheet_url, sheet = sheet_name)
glimpse(data)
print(colnames(data))
```

# Part 2: Store data in SQL database
### Changes
This function iterates through the rows and pulls each value from column in the sheet's data and assigns it into each value of the same column in MySQL database table:

```{r insert-data-into-SQL}
# Insert each row of data into MySQL table using parameterized queries
for (i in 1:nrow(data)) {
  query <- "INSERT IGNORE INTO Movies_new1 (Name, Gladiator2, Wolfs, The_Substance, Bad_Boys4, The_Beekeeper, Rebel_Ridge) VALUES (?, ?, ?, ?, ?, ?, ?)"
  dbExecute(con, query, params = list(
    data$Name[i],
    data$Gladiator2[i],
    data$Wolfs[i],
    data$The_Substance[i],
    data$Bad_Boys4[i],
    data$The_Beekeeper[i],
    data$Rebel_Ridge[i]
  ))
}
```

# Part 3: Transfer data from SQL database to R dataframe

This chunk transfers data from the SQL database into R dataframe `df`:

```{r transfer-data-from-SQL-to-R}
# Query to select all data from the table
query <- "SELECT * FROM Movies_new1"

# Execute the query and fetch the results
df <- dbGetQuery(con, query)

# Summary of df
glimpse(df)
summary(df)
```

# Part 4: Missing data strategy

Here I remove irrelevant `Timestamp` column and preview data:

```{r remove-Timestamp-column}
# Remove Timestamp column
df_zeroes_to_means <- df |> select(-Timestamp)
glimpse(df_zeroes_to_means)
```

This dataset contains some reviews with value of zero (0). These zero values are not actual reviews of the movies, but indicate that a person didn't watch this movie and didn't rate it in the survey.

This code replaces zero values in variables with the mean of the same variable:

```{r replace-zeroes-with-means}
# Replace zero values with column mean and round values
df_zeroes_to_means <- df_zeroes_to_means |>
  mutate(across(where(is.numeric), ~ {
    column_mean <- round(mean(.x[.x != 0], na.rm = TRUE))
    .x[.x == 0] <- column_mean
    round(.x, 0)
  }))

glimpse(df_zeroes_to_means)
summary(df_zeroes_to_means)
```

```{r save-data-to-CSV}
# Write the data to a CSV file
write.csv(df_zeroes_to_means, "C:/CUNY_MSDS/DATA607/LAB2/Movie_Reviews.csv", row.names = FALSE)

# Disconnecting from MySQL
dbDisconnect(con)
```

# Part 5: Conclusion
The summaries above show different statistical properties before and after replacing zeroes with means.

The reasons why I decided to replace missing values with means:

1. It can preserve statistical properties of the dataset.
2. It reduces bias and prevents skewing analysis results.
3. Variable with some missing values can affect the mean of that variable.
4. Data is more consistent and easier to interpret.
5. Missing and NA values can cause incorrect charts and graphs.
6. I didn't want to remove rows with zeroes, because it can cause loss of valuable information.

* * *
